Supplementary material

### 2021-11-02 analysis/paper/paper.Rmd
Should have rearranged radiocarbon dates correctly in radiocarbon.csv

# Data
All data that are mentioned in the text, and that are required to run the analyses are freely available in online repositories.

The data compiled for the analysis consists of all excavated Stone Age sites from the region between Horten county in the north-east, to Arendal in the south-west. These number a total nrow() sites, of which nrow() have radiocarbon dates. Of these, in turn, nrow() sites have radiocarbon dates to the Stone Age, and nrow() sites have radiocarbon dates to the Stone Age that were evaluated by the original excavators to relate to the lithic inventory on the sites (see also below). Spatial data in the form of features and site limits as defined by the excavation was retrieved from local databases at the MCH. Each radiocarbon date has been associated with site features or, where these weren't available, the spatial limit of the entire site. Due to somewhat variable practices between excavations, what spatial geometry best represents the site limit was decided based on an evaluation of the excavation reports. This means that the limits are variably given as area de-turfed before excavation, area stripped with excavator following the excavation, manually excavated area, or convex hulls generated around the site features. A note on this is given for each site in the supplemental material.

@@

The goal of this study is effectively to evaluate the spatial relationship between two phenomena, sites and shore-lines, each associated with their own range of temporal uncertainty. The first task was therefore to ascribe likely date ranges and a numerical expression of the associated uncertainty to the two phenomena.

Due to the dsfg there is a gradient in the rate of isostatic rebound, with increasing rebound rates towards the north-east. To take account of this gradient, the shoreline displacement curves were adjusted to each site location by taking into account the distance to the closest isobase. This was done by first finding the difference between the two closest displacement curves. The elevation difference was found for each year along the entirety of the curves, with each value divided by the meter distance between the isobases. This was done for both ends of the confidence intervals. To find the adjusted displacement curve for a given site location the distance between the site polygon and closest isobase was multiplied with the values for difference per meter. If the site is located south-west of the closest isobase, the elevation values were then subtracted from the original displacement curve, and if the site is located to the north-east the values were instead added to the original curve. The result of this process is shown for an example site in Figure.

@@

As the excavation of archaeological sites typically follow from residential and commercial development and the expansion of infrastructure, the area immediately surrounding the sites have sometimes been severely impacted by modern disturbances. In addition to employing 10 m resolution DTM to alleviate some of these issues, this also necessitated some additional editing of the elevation raster. This involved manually defining the extent of problem areas such as railways, highways, quarries and the like, using polygon features. The DTM values on these were then set to missing, and new elevation values were interpolated from the surrounding terrain. This was done using regularised spline interpolation with tension, using the default settings of r.fillnulls through GRASS GIS. In addition to being available in the , it has also been noted in the supplementary material when the area surrounding a site has been edited in this manner.

Armed with a likely date range for the occupation(s) of each site, the likely trajectory of relative sea-level change at that location, and an DTM edited to remove substantial modern , the simulation was performed. Each simulation run involved first drawing a single year from the posterior density estimate of a given occupation phase. This year then has a corresponding likely elevation range of the shoreline. An elevation value was drawn uniformly from this range, using intervals of 5 cm. The sea-level was then raised to this elevation on the DTM by defining all elevation values at or below this altitude as missing, and creating polygons of the resulting areas with missing values. The horisontal distance was then found by measuring the shortest distance between site and sea polygons, and the vertical distance by subtracting the. The so-called topographic distance between site and sea was also found by measuring the distance while taking into account the slope of the terrain on the DTM. This was done using the topoDist package. In the case where the sea-polygons intersects the site polygon, all distance measures were set to zero. In the case that the sea-polygons completely contain the site, the horisontal and topographic distance measures were made negative. While its safe to assume that an archaeological site was not occupied when it was located beneath sea-level, a negative result can reflect the inherent uncertainty in this procedure and might also help identify discrepancies in displacement data or radiocarbon dates.

>>>>

# Data
All data that are required to run the analyses and the derived data are freely available in an online repository. The dataset is slightly more extensive

The data compiled for the analysis consists of all excavated Stone Age sites from the region between Horten county in the north-east, to Arendal in the south-west. These number a total nrow() sites, of which nrow() have radiocarbon dates. Of these, in turn, nrow() sites have radiocarbon dates to the Stone Age, and nrow() sites have radiocarbon dates to the Stone Age that were evaluated by the original excavators to relate to the lithic inventory on the sites (see also below). Spatial data in the form of features and site limits as defined by the excavation was retrieved from local databases at the MCH. Each radiocarbon date has been associated with site features or, where these weren't available, the spatial limit of the entire site. Due to somewhat variable practices between excavations, what spatial geometry best represents the site limit was decided based on an evaluation of the excavation reports. This means that the limits are variably given as area de-turfed before excavation, area stripped with excavator following the excavation, manually excavated area, or convex hulls generated around the site features.

@@

The goal of this study is effectively to evaluate the spatial relationship between two phenomena, sites and shore-lines, each associated with their own range of temporal uncertainty. The first task was therefore to ascribe likely date ranges and associated uncertainty to the two phenomena.

Due to the dsfg there is a gradient in the rate of isostatic rebound, with increasing rebound rates towards the north-east. To take account of this gradient, the shoreline displacement curves were adjusted to each site location by taking into account the distance to the closest isobase. This was done by first finding the difference between the elevation values of the two closest displacement curves for each year along the entirety of the curves. Each value was then divided by the meter distance between the isobases. This was done for both ends of the confidence intervals. To find the adjusted displacement curve for a given site location the distance between the site polygon and closest isobase was multiplied with the values for difference per meter. If the site is located south-west of the closest isobase, the elevation values were then subtracted from the original displacement curve, and if the site is located to the north-east, the values were instead added to the original curve. The result of this process is shown for an example site in Figure.

@@

As the excavation of archaeological sites typically follow from residential and commercial development and the expansion of infrastructure, the area immediately surrounding the sites have sometimes been severely impacted by modern disturbances. In addition to employing 10 m resolution DTM to alleviate some of these issues, this also necessitated some additional editing of the elevation raster. This involved manually defining the extent of problem areas such as railways, highways, quarries and the like, using polygon features. The DTM values on these were then set to missing, and new elevation values were interpolated from the surrounding terrain. This was done using regularised spline interpolation with tension, using the default settings of r.fillnulls from GRASS GIS in R through the rgrass7 package. In addition to code and original spatial data being available in the research compendia for this paper, it has also been noted in the supplementary material when the area surrounding a site has been edited in this manner.

Armed with a likely date range for the occupation(s) of each site, the likely trajectory of relative sea-level change at that location, and an DTM edited to remove substantial modern, the simulation was performed. Each simulation run involved first drawing a single year from the posterior density estimate of a given occupation phase. This year then has a corresponding likely elevation range of the shoreline. An elevation value was drawn uniformly from this range, using intervals of 5 cm. The sea-level was then raised to this elevation on the DTM by defining all elevation values at or below this altitude as missing, and creating polygons of the resulting areas with missing values. The horisontal distance was then found by measuring the shortest distance between site and sea polygons, and the vertical distance by subtracting the elevation of the sea-level from the lowest elevation of the site polygon. The so-called topographic distance between site and sea was also found by measuring the distance while taking into account the slope of the terrain on the DTM. This was done using the topoDist package for R. In the case where the sea-polygons intersects the site polygon, all distance measures were set to zero. In the case that the sea-polygons completely contain the site, the horisontal and topographic distance measures were made negative. While its safe to assume that an archaeological site was not occupied when it was located beneath sea-level, a negative result can reflect the inherent uncertainty in this procedure and might also help identify discrepancies in displacement data or radiocarbon dates.

### 2021-11-02 analysis/supplementary-materials/supplementary.Rmd
Should have rearranged radiocarbon dates correctly in radiocarbon.csv

Subsequent rows for each site then show the simulation results. The first plot on these rows shows the probability density function from which site dates were drawn during simulation. The second plot shows the result of the simulation runs, where the intensity of the colour indicates the number of times the sea was simulated to be present at any given location. The third plot displays a violin plot of the measured distances between site and shore-line across all simulation runs. Measures for central tendency and dispersion associated with these are then given in the table

@@

Gunnarsrød 4 situated right by the railway (see also Gunnarsrød 5 below). The date is seen as possibly the result of old wood effect, a notion that is in line with simulation results here, as this date would the site located beneath the sea-level.

@@

Gunnarsrød 5 is today situated at the exit of a railway tunnel. The construction of the tunnel entrance has lead to a removal of parts of what used to be a hill on which the site was situated. While the interpolation appears to give reasonable results for the rest of the flatter marsh area where the railway runs today, the site itself would thus have been situated slightly higher, leading to the negative values on the distance measures. As the simulation results compared to the site map in   indicate that the site would not have been situated any significant distance from the shoreline, the values have therefore been set to zero for the final aggregative analysis.

>>>>

Subsequent rows for each site then show the simulation results. The first plot on these rows shows the probability density function from which site dates were drawn during simulation. The second plot shows the result of the simulation runs, where the intensity of the colour indicates the number of times the sea was simulated to be present at any given location. The third plot displays a violin plot of the measured distances between site and shore-line across all simulation runs. Measures for central tendency and dispersion associated with these are then given in the table
```{r anvik, include=FALSE}

```
Anvik is situated by the highway and a smaller road. Neither appear to significantly impact the simulation results. The dated fireplace is not seen as related to lithic material, which is believed to hail from the latest part of the Early Mesolithc (c. 8500-8250 BCE).

@@

Gunnarsrød 4 situated right by the railway (see also Gunnarsrød 5 below). The date was seen as possibly the result of old-wood effect, a notion that is in line with the simulation results here, as this date would have the site located beneath the sea-level.

@@

Gunnarsrød 5 is today situated at the exit of a railway tunnel. The construction of the tunnel entrance has lead to a removal of parts of what used to be a hill on which the site was situated. While the interpolation appears to give reasonable results for the rest of the flatter marsh area where the railway runs today, the site itself would thus have been situated slightly higher, likely leading to the negative values on the distance measures. As the simulation results compared to the site map in   indicate that the site would not have been situated any significant distance from the shoreline, the values have therefore been set to zero for the final aggregative analysis.
